A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-11-03 appendix.tex
adding kz=0 special case in response to inutsuka-san's query
more notes on kz=0 case (spurious?)

\section{One-fluid dispersion relation for the streaming instability}\label{compressible_streaming}
We consider an unstratified disk with $\Phi = \Phi (r)$ so that $\p_zP = \p_z\rho =
0$. The background $\p_rP/\rho$, $\tepsilon$ and $c_s^2$ are constant
input parameters. Then for the Epstein drag law, Eq. \ref{epstein}, we
have $\mathcal{C}=0$ in the background state. 
We Fourier analyze in $r$ and $z$ so that $\p_z\to \ii k_z$ and
$\p_r\to \ii k_x$ when acting on perturbations, and denote $|\bm{k}|^2
\equiv k_x^2 + k_z^2$. We neglect 
background gradients when compared with that of
perturbations. The linearized
equations are, after eliminating the azimuthal velocity:

@@

We eliminate the velocity perturbations to obtain
\begin{align}
  \left(\ii \zeta \sigma - \tstop c_s^2 \tepsilon |\bm{k}|^2\right)Q &=
  \ii \left[
  \frac{P}{\rho}\sigma + k_x\tstop c_s^2 F_r\left(1 -
  2\tepsilon\right)\right]W, \\
   \sigma^2 \left( \kappa^2 - \sigma^2 - \ii k_x F_r\right)W &=
    \left(k_z^2\kappa^2 - \sigma^2|\bm{k}|^2\right)Q, 
\end{align}
which yields the dispersion relation
\begin{align}
  &\frac{\zeta}{c_s^2|\bm{k}|^2}\sigma^5 + \ii \tepsilon \tstop
  \sigma^4 - \left[ \frac{\zeta}{c_s^2|\bm{k}|^2} \left(\kappa^2 - \ii
  k_x F_r\right) + \left(1 - \tepsilon\right)\right]\sigma^3 - \ii
  \tstop \left[\tepsilon \kappa^2 - \ii k_x F_r\left( 1 -
  \tepsilon\right)\right]\sigma^2 \notag\\ 
  &+ \left(1 -
  \tepsilon\right)\left(\frac{k_z\kappa}{|\bm{k}|}\right)^2\sigma +
  k_x \tstop F_r\left(1 - 2\tepsilon\right)
  \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2  = 0.\label{streaming_dispersion}
\end{align}
The equation of state $P = c_s^2\left( 1 - \tepsilon\right)\rho$
was used. If we now set $\zeta = 0$ or consider $c_s^2\to \infty$ to
obtain the 
incompressible limit, and define $\tau_\mathrm{s} \equiv
\tstop/\left(1 - \tepsilon\right)$, we obtain: 
\begin{align*}
\ii \tepsilon \tau_\mathrm{s}
  \sigma^4 - \sigma^3 - 
  \tau_\mathrm{s} \left[ \ii \tepsilon \kappa^2 + k_x F_r\left( 1 -
  \tepsilon\right)\right]\sigma^2 
  + \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2\sigma + 
  k_x \tau_\mathrm{s} F_r\left(1 - 2\tepsilon\right)
  \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2  = 0,
\end{align*}
as derived by \cite{jacquet11} and \cite{laibe14} for the streaming
instability. If $|\sigma|/\OmK = O(\tau_\mathrm{s}\OmK)$ and
$\tau_\mathrm{s}\OmK\ll 1$, then the quartic term
is small and may be neglected. In that case we obtain the cubic
dispersion relation of \citet{youdin05a}.

>>>>

\section{One-fluid dispersion relation for the streaming instability}\label{compressible_streaming}
We consider an unstratified disk with $\Phi = \Phi (r)$ so that $\p_zP = \p_z\rho =
0$. The background $\p_rP/\rho$, $\tepsilon$ are constant
input parameters. Then for the Epstein drag law, Eq. \ref{epstein}, we
have $\mathcal{C}=0$ in the background state. 
We Fourier analyze in $r$ and $z$ so that $\p_z\to \ii k_z$ and
$\p_r\to \ii k_x$ when acting on perturbations, and denote $|\bm{k}|^2
\equiv k_x^2 + k_z^2$. We consider large $k_x$ and 
%background gradients when compared with that of 
%perturbations. 
neglect the radial density gradient compared to that of pertuburbations 
(formally setting $\p_r\rho=0$). This is also done in most local studies of
dusty disks \citep[e.g.][]{youdin07b}.  
The linearized
equations are, after eliminating the azimuthal velocity:

@@

We eliminate the velocity perturbations to obtain
\begin{align}
  \left(\ii \zeta \sigma - \tstop c_s^2 \tepsilon |\bm{k}|^2\right)Q &=
  \ii \left[
  \frac{P}{\rho}\sigma + k_x\tstop c_s^2 F_r\left(1 -
  2\tepsilon\right)\right]W, \\
   \sigma^2 \left( \kappa^2 - \sigma^2 - \ii k_x F_r\right)W &=
    \left(k_z^2\kappa^2 - \sigma^2|\bm{k}|^2\right)Q, 
\end{align}
which yields the dispersion relation
\begin{align}
  &\frac{\zeta}{c_s^2|\bm{k}|^2}\sigma^5 + \ii \tepsilon \tstop
  \sigma^4 - \left[ \frac{\zeta}{c_s^2|\bm{k}|^2} \left(\kappa^2 - \ii
  k_x F_r\right) + \left(1 - \tepsilon\right)\right]\sigma^3 - \ii
  \tstop \left[\tepsilon \kappa^2 - \ii k_x F_r\left( 1 -
  \tepsilon\right)\right]\sigma^2 \notag\\ 
  &+ \left(1 -
  \tepsilon\right)\left(\frac{k_z\kappa}{|\bm{k}|}\right)^2\sigma +
  k_x \tstop F_r\left(1 - 2\tepsilon\right)
  \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2  = 0.\label{streaming_dispersion}
\end{align}
The equation of state $P = c_s^2\left( 1 - \tepsilon\right)\rho$
was used. If we now set $\zeta = 0$ or consider $c_s^2\to \infty$ to
obtain the 
incompressible limit, and define $\tau_\mathrm{s} \equiv
\tstop/\left(1 - \tepsilon\right)$, we obtain: 
\begin{align}
\ii \tepsilon \tau_\mathrm{s}
  \sigma^4 - \sigma^3 - 
  \tau_\mathrm{s} \left[ \ii \tepsilon \kappa^2 + k_x F_r\left( 1 -
  \tepsilon\right)\right]\sigma^2 
  + \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2\sigma + 
  k_x \tau_\mathrm{s} F_r\left(1 - 2\tepsilon\right)
  \left(\frac{k_z\kappa}{|\bm{k}|}\right)^2  = 0,\label{streaming_incompressible}
\end{align}
as derived by \cite{jacquet11} and \cite{laibe14} for the streaming
instability. If $|\sigma|/\OmK = O(\tau_\mathrm{s}\OmK)$ and
$\tau_\mathrm{s}\OmK\ll 1$, then the quartic term
is small and may be neglected. In that case we obtain the cubic
dispersion relation of \citet{youdin05a}.

### 2016-11-03 linear.tex
more notes on kz=0 case (spurious?)

Eq. \ref{lin_mass}---\ref{lin_energy} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained. We now discuss solutions to these equations in an unstratified disk 
 for the streaming instability (\S\ref{si}), and a stratified disk for the VSI (\S\ref{results}).

@@

\subsection{Overstable epicycles}
If we further consider the limit $k_z=0$, the incompressible dispersion
relation, Eq. \ref{streaming_incompressible}, reduces to 
\begin{align}      %streaming_incompressible
\ii \sigma^2 - \frac{\sigma}{\epsilon \tstop}  - \left(\ii \kappa^2 +
  \frac{k_xF_r}{\epsilon}\right) = 0. \label{kz_zero_si}
\end{align} 
%Here we use $\tau_\mathrm{s}=\tstop/(1-\tepsilon)$ for convenience. 
If the quadratic term is neglected, then $\imag{\sigma}<0$, implying
stability, as concluded by \cite{youdin05a}.

Now, writing $\sigma = \ii s - \omega$, the real and imaginary parts of 
Eq. \ref{kz_zero_si}  are
\begin{align}
\omega^2 - s^2 - \frac{s}{\epsilon\tstop}  - \kappa^2 &= 0, \label{epi_overs1}\\
\omega\left(2 s + \frac{1}{\epsilon\tstop}\right) - \frac{k_xF_r}{\epsilon} & =
                                                                     0.  \label{epi_overs2}
\end{align}
The growth rate satisfies
\begin{align*}
\left( \frac{k_xF_r\tstop}{1 + 2 s\epsilon\tstop}\right)^2 - s^2
  - \frac{s}{\epsilon \tstop} = \kappa^2. %\omega = \frac{k_xF_r\tstop}{1 + 2 s\epsilon\tstop}
\end{align*}

Assuming slow growth such that $\epsilon s \tstop \ll 1$, we find 
\begin{align}
  s \simeq \epsilon \tstop \left(k_xF_r\tstop\right)^2 - \kappa^2,
\end{align}
Growth is possible for
sufficiently large stopping times, and/or large radial pressure
gradients, and/or sufficiently small radial wavelengths. 
%For small growth rates the real frequency $|\omega| =  
%O(\kappa)$, so these are overstable epicycles.

It is interesting to note that Eq. \ref{epi_overs1}---\ref{epi_overs2}
admit pure oscillations with $s \equiv 0$, $\omega = 
\pm\kappa$ if 
\begin{align}
  \tstop = \frac{\kappa}{k_x\left|F_r\right|}. 
\end{align}
This is the condition for marginal stability.

@@

\citep[cf. the `convective overstability', ][]{klahr14,lyra14,latter16}.

@@

{\bf note: important to use largrangian pressure pert properly}
We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\rho\left(\left|\dd
  v_x\right|^2+\left|\dd
  v_z\right|^2\right)}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. %, or directly from
%Eq. \ref{streaming_mass}---\ref{streaming_vz}. 
Interestingly, we find that for cases with
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta
P\simeq \ii\dd v_x \p_rP/\sigma$, i.e. the radial pressure gradient
is responsible for growth. Conversely, for `linB' with $\epsilon
< 1$, one can approximate $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. This
suggests that for SI in gas-dominated disks, the distinction between
Eulerian and Lagragian pressure perturbations is unimportant. 
{\bf but weak growth in this case}
%However
%in that case the gro

@@

{\bf suppose large scale pressure grad drops outwards. 
pert the sys by kicking more dust inwards. gas moves out, into region 
  of lower pressure in the bg. i.e. create pressure bump at larger
  radius. it attracts the first dust particles back out, plus some more
  because bump is larger. (move gas from higher density to lower
  density produce asymmetric bump/trough.}

@@

To demonstrate the above instabilities we numerically solve the
full dispersion relation, Eq. \ref{streaming_dispersion} at fixed $K_x
= 1500$ and $\epsilon=2$. Growth rates are shown in 
Fig. \ref{dusty_growth} as a function of $K_z$ at fixed $\taus\OmK=0.01$
(top panel, which includes `linC' in Table \ref{si_compare}); and as a
function of $\taus\OmK$ at fixed $K_z=50$ (bottom panel).

The top panel shows that the streaming instability vanishes as
$K_z\to0$, but it dominates most of $K_z$
-space. The overstable epicycles become important for 
$K_z\lesssim 100$, but their growth rates are weak. We thus
expect streaming instabilities to dominate, provided small
lengthscales can be resolved.

In the bottom panel we confirm that the overstable epicycles require a
sufficiently large stopping time at fixed wavenumbers. The vertical
dashed line is the critical stopping time 
\begin{align*}
  \tau_{\mathrm{s}0}\OmK = \frac{1}{2K_x\left(1-\tepsilon\right)^2} 
\end{align*}
in the above notation. For this example, $\tau_{\mathrm{s}0}\OmK\simeq
0.003$, as observed. On the other hand, the streaming instability
shows weak dependence on $\taus$ for this range of stopping times.

\begin{align*}
  s = \epsilon \left(1 - \tepsilon
  \right)\taus\OmK^2\left[4K_x^2\left(1-\tepsilon\right)^4\left(\taus\OmK\right)^2
  -1 \right]
\end{align*}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/streaming2}
  \includegraphics[width=\linewidth]{figures/streaming3}
  \caption{Growth rates of dust-drag instabilities with fixed radial
    wavenumber and dust-to-gas ratio, as a function of
    the vertical wavenumber at fix stopping time (top); and as a
    function of stopping time at fixed vertical wavenumber (bottom).  
\label{dusty_growth}
  }
\end{figure}

>>>>

Eq. \ref{lin_mass}---\ref{lin_energy} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained. We now discuss solutions to these equations in an unstratified disk 
in \S\ref{si} and a stratified disk in \S\ref{results}.

@@

\subsection{Overstable dusty epicycles}
If we further consider the limit $k_z=0$, the incompressible dispersion
relation, Eq. \ref{streaming_incompressible}, reduces to 
\begin{align}      %streaming_incompressible
\ii \sigma^2 - \frac{\sigma}{\epsilon \tstop}  - \left(\ii \kappa^2 +
  \frac{k_xF_r}{\epsilon}\right) = 0. \label{kz_zero_si}
\end{align} 
%Here we use $\tau_\mathrm{s}=\tstop/(1-\tepsilon)$ for convenience. 
Neglecting the quadratic term leads to stability, $\imag{\sigma}<0$,
as concluded by \cite{youdin05a}.

However, the real and imaginary parts of Eq. \ref{kz_zero_si}  are
\begin{align}
\omega\left(2 s + \frac{1}{\epsilon\tstop}\right) - \frac{k_xF_r}{\epsilon} & =
                                                                     0, \label{epi_overs1} \\
\omega^2 - s^2 - \frac{s}{\epsilon\tstop}  - \kappa^2 &= 0, \label{epi_overs2}
\end{align}
so the growth rate satisfies 
\begin{align*}
\left( \frac{k_xF_r\tstop}{1 + 2 s\epsilon\tstop}\right)^2 - s^2
  - \frac{s}{\epsilon \tstop} = \kappa^2. %\omega = \frac{k_xF_r\tstop}{1 + 2 s\epsilon\tstop}
\end{align*}

Assuming slow growth such that $\epsilon s \tstop \ll 1$, we find 
\begin{align}
  &|\omega| \simeq \left|k_xF_r\right| \tstop, \\
  &s \simeq \epsilon \tstop \left[\left(k_xF_r\tstop\right)^2 - \kappa^2\right],\label{epi_grow}
\end{align}
Notice the $\tstop$ and $\tstop^3$ dependence for 
$\omega$ and $s$. This is also true for the streaming 
instability for sufficiently small $\tstop$ \citep{jacquet11}. Unlike
SI though, here there can be growth when $\epsilon=1$, whereas the SI
is severly suppressed at equal dust-to-gas ratio.

According to Eq. \ref{epi_grow}, growth is possible for sufficiently
large stopping times, and/or large radial pressure gradients, and/or
sufficiently small radial wavelengths. For example, given $k_xF_r$
instability requires 
\begin{align}
\tstop > \frac{\kappa}{k_x\left|F_r\right|}\equiv t_{\mathrm{s}0}.\label{epi_grow_criterion}
\end{align}
The mode is marginally stable if $\tstop = t_{\mathrm{s}0}$.

These slowly-growing modes have $|\omega|\simeq \kappa$, so they correspond to
overstable epicycles. This property is similar to the `convective 
overstability' discussed in dust-free disks 
\citep{klahr14,lyra14,latter16}. In that case, the combination of an 
unstable radial entropy gradient ($N_r^2  \propto \p_rP\p_rS < 0 $) and
finite thermal relaxation time leads to growing epicycles. Indeed, the
effective radial buoyancy frequency in our dusty disk 
with negligble background density gradient is unstable, 
\begin{align}
  N_r^2 \equiv F_r\frac{\p S_\mathrm{eff}}{\p r} =
  -\frac{\rho}{P}F_r^2<0.  
\end{align}

Convective overstability cannot exist in our locally isothermal disks,
which have zero cooling times. On the other hand, dust-gas drag 
is equivalent to heating/cooling in our framework. Thus it may not be 
surprising to find a dusty analog of the convective 
overstability. Both growth rates scale with 
$|N_r^2|$, and both require finite effective cooling times.

However, the structure of the modes differ. The dusty epicycles 
have no vertical structure and require $k_x\neq0$; whereas the
convective overstability has no radial structure but require 
$k_z\neq0$. Dusty epicycles may thus operate in  
razor-thin disks, but convective overstability requires the vertical
dimension.

An important caveat is that the one-fluid approximation for dusty gas 
may not hold well when $|\omega| \gtrsim 
\Omega$ \citep{youdin05a}, as is the case for dusty epicycles. Their
existence should be checked with full dust-plus-gas 
simulations. Several 2D, razor-thin disk simulations do suggest a
dusty instability that destory vortices
\citep{fu14b,raettig15,surville16}. These setups are significantly
more complex than our models, but \citeauthor{raettig15} did find 
particle clumping in a test simulation without a vortex. These results
might be related to the overstable dusty epicycles discussed here.

%While there is some uncertainty on the reality of the overstable dusty
%epicyclic modes, 
Real disks are three-dimensional, so the importance of dusty 
epicycles should be compared with $k_z\neq0$ modes, which
includes the classic streaming instability \citep{youdin05a},
discussed next.

%For small growth rates the real frequency $|\omega| =  
%O(\kappa)$, so these are overstable epicycles.

%We finally note that Eq. \ref{epi_overs1}---\ref{epi_overs2} 
%admit pure oscillations with $s \equiv 0$, $\omega = 
%\pm\kappa$ if 
%\begin{align}
%  \tstop = \frac{\kappa}{k_x\left|F_r\right|}. 
%\end{align}
%This is the condition for marginal stability.

@@

@@

{\bf note: important to use largrangian pressure pert properly}

We also checked the growth rates $s$ satisfy 
\begin{align} 
  s = \frac{\left|\sigma\right|^2\imag\left(\Delta P
    \Delta\rho^*/\rho\right)}{2\real\left(\sigma\right)\rho\left(\left|\dd
  v_x\right|^2+\left|\dd
  v_z\right|^2\right)}, \label{si_check}
\end{align}
as implied by Eq. \ref{thermal_instability} and
Eq. \ref{pdv}. %, or directly from
%Eq. \ref{streaming_mass}---\ref{streaming_vz}. 
Interestingly, we find that for cases with
$\epsilon>1$, Eq. \ref{si_check} can be satisified with $\Delta
P\simeq \ii\dd v_x \p_rP/\sigma$, i.e. the radial pressure gradient
is responsible for growth. Conversely, for `linB' with $\epsilon
< 1$, one can approximate $\Delta P \simeq \dd P$ in Eq. \ref{si_check}. This
suggests that for SI in gas-dominated disks, the distinction between
Eulerian and Lagragian pressure perturbations is unimportant. 
{\bf but weak growth in this case}
%However
%in that case the gro

@@

%{\bf suppose large scale pressure grad drops outwards. 
%pert the sys by kicking more dust inwards. gas moves out, into region 
%  of lower pressure in the bg. i.e. create pressure bump at larger
%  radius. it attracts the first dust particles back out, plus some more
%  because bump is larger. (move gas from higher density to lower
%  density produce asymmetric bump/trough.}

@@

We numerically solve the
full dispersion relation, Eq. \ref{streaming_dispersion} at fixed $K_x
= 1500$ and $\epsilon=2$. Growth rates are shown in 
Fig. \ref{dusty_growth1} as a function of $K_z$ at fixed $\taus\OmK=0.01$
(which includes `linC' in Table \ref{si_compare}); and in
Fig. \ref{dusty_growth2} as a function of $\taus\OmK$ at fixed
$K_z=50$.

Fig. \ref{dusty_growth1} show the overstable dusty epicycles' growth
rates are weakly depedent on $K_z$. They are well approximated by
that in the $K_z=0$ limit, Eq. \ref{epi_grow}, which becomes 
\begin{align*}
  s = \epsilon \left(1 - \tepsilon
  \right)\taus\OmK^2\left[4K_x^2\left(1-\tepsilon\right)^4\left(\taus\OmK\right)^2
  -1 \right]
\end{align*}
in the above notation. 
For the case considerd in Fig. \ref{dusty_growth1} we find
$s\simeq0.067\OmK$, as observed. By contrast, the streaming
instability requires $K_z>0$, and dominates when
$K_z\gtrsim 100$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/streaming2}
  \caption{Growth rates of dust-drag instabilities with fixed radial
    wavenumber and dust-to-gas ratio, as a function of
    the vertical wavenumber at fix stopping time. \label{dusty_growth1}
  }
\end{figure}

In Fig. \ref{dusty_growth2} we fix the wavenumbers and vary the
stopping time. The vertical dashed line is the critical stopping time
for the overstable dusty epicycles, Eq. \ref{epi_grow_criterion},
which is  
\begin{align*}
  \tau_{\mathrm{s,min}}\OmK = \frac{1}{2K_x\left(1-\tepsilon\right)^2} 
\end{align*}
in the above notation. For this example, $\tau_{\mathrm{s,min}}\OmK\simeq 
0.003$, as observed. On the other hand, the streaming instability
shows weak dependence on $\taus$ for this range of stopping times 
(it ultimated vanishes as $\taus\to0$).

\begin{figure}
  \includegraphics[width=\linewidth]{figures/streaming3}
  \caption{Growth rates of dust-drag instabilities with fixed radial
    wavenumber and dust-to-gas ratio, as a
    function of stopping time at fixed vertical wavenumber. The
    vertical dashed line correspond to the minimum stopping time for
    growth of the dusty epicycle, given by
    Eq. \protect\ref{epi_grow_criterion}. \label{dusty_growth2}
  }
\end{figure}

These calculations suggest that in a realistic 3D dusty disk where
vertical motions are allowed, the classic streaming
instability is dominant because there exists $K_z$ values for which it
dominates. This also renders the uncertainties in the  
approximations used to describe the dusty epicycle modes
irrelevant. On the other hand, the one-fluid framework can capture the
streaming instability in the appropriate limit, as seen in Table
\ref{si_compare}.
